Grouping data and nesting

I also don't think we should consider the protein column. Is the amount of protein_groups the same as the amount of genes? If we subset the data again, by grouping for cell_type, what is then the lenght of them? How do I figure out how many unique genes there are?

{r}
df_no_na_long |> 
  group_by(cell_type) |> 
  summarize(n = n())
#We can see that there are equally many cmp, gmp, hsc and mep but there are more clp's which is due to the no of replicates

{r}
df_no_na_long |> 
  ungroup() |> 
  select(genes) |> 
  distinct()

#There are 3623 different genes
df_no_na_long |> 
  ungroup() |> 
  select(protein_groups) |> 
  distinct()
#There are 3724 different protein groups 
#why is the number of genes and number of proteins not the same??
#And how do we figure out if the same genes are present in all of the cell types?


{r}
df_no_na_long |> 
  ungroup() |> 
  select(genes, cell_type) |> 
  group_by(cell_type) |> 
  summarize(n = n())

{r}
df_no_na_long |> 
  filter(cell_type == "hsc" | cell_type == "gmp") |> 
  distinct()

Now lets try to work with hsc and gmp.

I want to create a new variable that shows me the ration between hsc and gmp.

{r}
df_no_na_long_sample |> 
  ungroup() |> 
  select(cell_type, replicate_n, genes, intensity) |> 
  filter(cell_type == "hsc" | cell_type == "gmp") |> 
  group_by(cell_type) |> 
  mutate(ratio = case_when(cell_type == "hsc" ~ 1,
                           cell_type == "gmp" ~ 0))

{r}
this <- df_no_na_long |> 
  ungroup() |> 
  select(cell_type, replicate_n, genes, intensity) |> 
  pivot_wider(names_from = cell_type,
              values_from = intensity) |> 
  filter(replicate_n != 5) |> 
  group_by(genes) |> 
  nest() |> 
  ungroup()

{r}
this |> 
  filter(genes == "Gprin1") |> 
  pull(data)
  
  
  
  And then from CMP to MEP and from CMP to GMP. Because that is the order in which the cells develop.

```{r}
library(ggrepel) 
test_plot |>
  #When are points significant? In this case if fold > 1 it is overexpressed
 # mutate(label = case_when(fold > 1 ~ protein_groups, fold < 1 ~ "")) |>
  ggplot(aes(x = log2fold, y = desc(log10(pval)), color = is_signif))+
  geom_point(size=1, alpha = 0.5)+
#  geom_text_repel(aes(label = label), size = 3)+
  theme_minimal()+
  labs(title = "Expression of CLP relative to HSC",
       x = "fold",
       y = "-log10(p)",
       caption = "proteomics data")+
  theme(legend.position = "none")+
  #The horizontal line that indicates significance - remember to check if it correct with qvalue and all
#  geom_hline(yintercept = 2)+
  scale_color_manual(values = c("overexpressed" = "#D55E00", "underexpressed" = "#0072B2", "not_significant" = "gray"))

#The threshold should be p value of 0.05
#Use q-value instead of pvalue 
#Sofias code for defining the color
#scale_color_manual(values = c("Down-regulated" = "blue", "Up-regulated" = "red", "NA" = "gray"))
```

```{r}
hsc_clp_plot <- df_volcano_clp_hsc |>
  ggplot(aes(x = fold_log2, y = desc(log10(q_val)), color = expression))+
  geom_point(size=0.5, alpha = 0.5)+
  theme_classic()+
  labs(title = "Expression of CLP relative to HSC",
       x = "Log2fold-change",
       y = "-log10(q)",
       caption = "proteomics data") +
  theme(legend.position = "none") +
  #The horizontal line that indicates significance - remember to check if it correct with qvalue and all
#  geom_hline(yintercept = 2)+
  scale_color_manual(values = c("overexpressed" = "#D55E00", "underexpressed" = "#0072B2", "not_significant" = "gray"))
```

```{r}
hsc_cmp_plot <- df_volcano_cmp_hsc |>
  ggplot(aes(x = fold_log2, y = desc(log10(q_val)), color = expression))+
  geom_point(size=0.5, alpha = 0.5)+
  theme_classic()+
  labs(title = "Expression of CMP relative to HSC",
       x = "Log2fold-change",
       y = "-log10(q)",
       caption = "proteomics data")+
  theme(legend.position = "none")+
  #The horizontal line that indicates significance - remember to check if it correct with qvalue and all
#  geom_hline(yintercept = 2)+
  #Colors are from the colorblind palette
  scale_color_manual(values = c("overexpressed" = "#D55E00", "underexpressed" = "#0072B2", "not_significant" = "gray"))
```

```{r}
cmp_mep_plot <- df_volcano_mep_cmp |>
  ggplot(aes(x = fold_log2, y = desc(log10(q_val)), color = expression))+
  geom_point(size=0.5, alpha = 0.5)+
  theme_classic()+
  labs(title = "Expression of MEP relative to CMP",
       x = "Log2fold-change",
       y = "-log10(q)",
       caption = "proteomics data")+
  theme(legend.position = "none")+
  #The horizontal line that indicates significance - remember to check if it correct with qvalue and all
#  geom_hline(yintercept = 2)+
scale_color_manual(values = c("overexpressed" = "#D55E00", "underexpressed" = "#0072B2", "not_significant" = "gray"))
```

```{r}
cmp_gmp_plot <- df_volcano_gmp_cmp |>
  ggplot(aes(x = fold_log2, y = desc(log10(q_val)), color = expression))+
  geom_point(size=0.5, alpha = 0.5)+
  theme_classic()+
  labs(title = "Expression of GMP relative to CMP",
       x = "Log2fold-change",
       y = "-log10(q)",
       caption = "proteomics data")+
  theme()+
  #The horizontal line that indicates significance - remember to check if it correct with qvalue and all
#  geom_hline(yintercept = 2)+
  scale_color_manual(values = c("overexpressed" = "#D55E00", "underexpressed" = "#0072B2", "not_significant" = "gray"))
```
Therefore we will create four volcanoplots that compare the different cell-type (stages). CLP will be compared to HSC. CMP will be comared to HSC. And then MEP and GMP will be compared to CMP. In order to prepare the data for these volcanoplots there must be for each pair of compared cell-types be created a variable of log2-fold-change and of q-value. To figure out this data wrangling took quite a bit of effort. In the end what was done is to subset the full dataframe to the two cell types in question, then summarize to mean and standard deviation values, then transfer to wide format, and then compute the ratio of means for each protein group as well as compute the p-value which is then p.adjusted to the q-value. In order to compute p-value based on means and standard deviations a function was created that computes p-value for a two-sided student t-test of two populations based on means, standard deviations and number of replicates as input.

## CLP relative to HSC

```{r}
#starting form df_no_na_long_1 (will need to change the tibble when the universal is done)
df_hsc_clp <- df_no_na_long_1 |> 
  group_by(protein_groups,cell_type) |> 
    #subsetting the data
  filter(cell_type == c("hsc", "clp")) |> 
 #  take the mean/sd of of the replicates for the protein groups and cell types.
   summarise(
    mean = mean(intensity),
    sd = sd(intensity)) |> 
  #transform to wide format
      pivot_wider(names_from = cell_type, values_from = c(mean, sd))

#Computing new variables that calculate fold, p_val and q_val

df_hsc_clp <- df_hsc_clp |> 
  mutate(fold_log2 = log2(mean_clp/mean_hsc), 
         #See the p-value function in 99_functions.R
       p_val = pval(mu1 = mean_clp, mu2 = mean_hsc, n1 = 5, n2 = 4, s1 = sd_clp, s2 = sd_hsc),
       q_val = (p.adjust(p_val)))

#including variable overexpressed stating if proteins are over- or underexpressed or not significant
df_hsc_clp <- df_hsc_clp |> 
  mutate(expression = case_when(fold_log2 > 0 & q_val <= 0.05~ "overexpressed",
                                fold_log2 < 0 & q_val <= 0.05 ~ "underexpressed",
                                q_val >0.05  ~ "not_significant"))

#removing colomns unnecesary for volcanoplot
df_hsc_clp <- df_hsc_clp |> 
  select(protein_groups, fold_log2, q_val, expression)
```

Repeating the same data wrangling for the remaining three volcano plots is automatized using the function volcano_augment which can be seen in 99_functions.R.

## CMP relative to HSC

```{r}
#Right now we have the issue that the automatized function does not produce the same output as the original code... Now the question is which one of them is right and which one is wrong..? 
df_new_cmp_hsc <- volcano_augment(df = df_no_na_long_1, later_cell = "cmp", earlier_cell = "hsc")
#Remember to delete the one for hsc as that one is supposedly already created in the code above. And remember to delete new for them all. ANd figure out which of the functions is the correct one!!
df_new_clp_hsc <- volcano_augment(df = df_no_na_long_1, later_cell = "clp", earlier_cell = "hsc")
df_new_mep_cmp <- volcano_augment(df = df_no_na_long_1, later_cell = "mep", earlier_cell = "cmp")
df_new_gmp_cmp <- volcano_augment(df = df_no_na_long_1, later_cell = "gmp", earlier_cell = "cmp")

```

```{r}
df_hsc_cmp <- df_no_na_long_1 |> 
  group_by(protein_groups,cell_type) |> 
  filter(cell_type == c("hsc", "cmp")) |> 
   summarise(
    mean = mean(intensity),
    sd = sd(intensity)) |> 
      pivot_wider(names_from = cell_type, values_from = c(mean, sd)) |> 
  mutate(fold_log2 = log2(mean_cmp/mean_hsc), 
       p_val = pval(mean_cmp, mean_hsc, 4, 4, sd_cmp, sd_hsc),
       q_val = (p.adjust(p_val))) |> 
  mutate(expression = case_when(fold_log2 > 0 & q_val <= 0.05~ "overexpressed",
                                fold_log2 < 0 & q_val <= 0.05 ~ "underexpressed",
                                q_val >0.05  ~ "not_significant")) |> 
  select(protein_groups, fold_log2, q_val, expression)
```

## MEP relative to CMP

```{r}
df_cmp_mep <- df_no_na_long_1 |> 
  group_by(protein_groups,cell_type) |> 
  filter(cell_type == c("mep", "cmp")) |> 
   summarise(
    mean = mean(intensity),
    sd = sd(intensity)) |> 
      pivot_wider(names_from = cell_type, values_from = c(mean, sd)) |> 
  mutate(fold_log2 = log2(mean_cmp/mean_mep), 
       p_val = pval(mean_mep, mean_cmp, 4, 4, sd_mep, sd_cmp),
       q_val = (p.adjust(p_val))) |> 
  mutate(expression = case_when(fold_log2 > 0 & q_val <= 0.05~ "overexpressed",
                                fold_log2 < 0 & q_val <= 0.05 ~ "underexpressed",
                                q_val >0.05  ~ "not_significant")) |> 
  select(protein_groups, fold_log2, q_val, expression)
```

## GMP relative to CMP

```{r}
df_cmp_gmp <- df_no_na_long_1 |> 
  group_by(protein_groups,cell_type) |> 
  filter(cell_type == c("gmp", "cmp")) |> 
   summarise(
    mean = mean(intensity),
    sd = sd(intensity)) |> 
      pivot_wider(names_from = cell_type, values_from = c(mean, sd)) |> 
  mutate(fold_log2 = log2(mean_cmp/mean_gmp), 
       p_val = pval(mean_gmp, mean_cmp, 4, 4, sd_gmp, sd_cmp),
       q_val = (p.adjust(p_val))) |> 
  mutate(expression = case_when(fold_log2 > 0 & q_val <= 0.05~ "overexpressed",
                                fold_log2 < 0 & q_val <= 0.05 ~ "underexpressed",
                                q_val >0.05  ~ "not_significant"))
#  select(protein_groups, fold_log2, q_val, expression)
```

First of all we want to subset the data so we only work with two cell types. In this case hsc and gmp as they appear to be the most different.

```{r}
#Due to an error occuring when saving ggplot directly in the pipeline of the latest plot which apparently is caused by a recent edit in ggplot2-3.3.4 the saving of plots is done separately.
ggsave(plot = hsc_clp_plot, filename = "hsc_clp_volcano_plot.png", width = 20, height = 10, units = "cm", path = here("results"))
ggsave(plot = hsc_cmp_plot, filename = "hsc_cmp_volcano_plot.png", width = 20, height = 10, units = "cm", path = here("results"))
ggsave(plot = cmp_mep_plot, filename = "cmp_mep_volcano_plot.png", width = 20, height = 10, units = "cm", path = here("results"))
ggsave(plot = cmp_gmp_plot, filename = "cmp_gmp_volcano_plot.png", width = 20, height = 10, units = "cm", path = here("results"))
```

A future direction for the project would be to assign all the protein groups categories based on their functionality. It would for instance be interesting to see which are the proteins that are overexpressed in the different cell types, but since there are so many of them it does not make sense to label all of them. We might for instance label the proteins that have a 2logfold change above 2, which means they are overexpressed by 400%.

Lets have a look at hsc to cmp. We see that three protein groups are significantly overexpressed above 400%. P68033, P05213 and Q9WUA1. We can look up their Uniprot acession ID's and we find the following information:

P68033 is an actin, actins are highly conserved proteins that are involved in various types of cell motility and are ubiquitously expressed in all eukaryotic cells.

```{r}
# Histogram that shows distribution of the fold
ggplot(df_mean_intensitities, aes(x=fold))+
  geom_histogram()+
  scale_x_continuous(limits = c(0.75, 1.25))
```

### Toyplot

```{r}
#This is a toyplot to try out new things
df_new_hsc_cmp|>
  mutate(label = case_when(expression == "overexpressed" & fold_log2 > 2 ~ protein_groups,
                           .default = "")) |> 
  ggplot(aes(x = fold_log2, y = desc(log10(q_val)), color = expression))+
  geom_point(size=0.5, alpha = 0.5)+
  theme_classic()+
  geom_text_repel(aes(label = label), size = 3)+
  labs(title = "Expression of GMP relative to CMP",
       x = "Log2fold-change",
       y = "-log10(q)",
       caption = "proteomics data")+
  theme(legend.position = "bottom", legend.title = element_blank())+
  guides(color = guide_legend(overrisde.aes = list(size = 1000)))+
  #The horizontal line that indicates significance - remember to check if it correct with qvalue and all
#  geom_hline(yintercept = 2)+
  scale_color_manual(values = c("overexpressed" = "#D55E00", "underexpressed" = "#0072B2", "not_significant" = "gray"))
  
```