
## Make PCA plot

```{r}
library(tidyr)
library('ggbiplot')

## Libraries

```{r 01_libraries_and_options}
library(tidyr)

library(graphics)

# Set the bitmap type to 'cairo' to avoid PNG issue
options(bitmapType = 'cairo')


#Input data tois wide format, excluding columns with missing data.


# The new data-frame for df_no_na_wide is called : df_feature_wide.
# For creating a column sample as a guiding column use:
# mutate

df_feature_wide |> 
  mutate(sample = str_c(cell_type, replicate_n, sep = "_"))


df_input <- df_no_na_wide |> select(-c('sample','cell_type')) |> 
  mutate(across(where(is.list), ~ sapply(., function(x) x[1]))) |> 
  na.omit() 

colourby <- df_no_na_wide$cell_type

```

## PCA plots and variance analysis

```{r 01_principal_components}
df_feature_wide |> 
  mutate(sample = str_c(cell_type, replicate_n, sep = "_"))

colourby <- df_feature_wide$cell_type

#Projecting data unto principal components (pc)
pc <- prcomp(df_input, center = FALSE, scale = FALSE)


#Gives overview of variability of PC
#print(pc)

#Gives an overview, you should be able to tell the variance already here even without the plot
summary(pc)

#Takes the principal component values (pc$x)  

#Gives an overview, you should be able to tell the variance already here even without the plot
summary(pc)
```

```{r 03_pca_plot}
#Plots the principal component values (pc$x)  


pc$x |> as.data.frame() |>
  ggplot(aes(x = PC1, y = PC2, color = colourby)) +
  geom_point(size = 2.5)
```


## Save plot

```{r}

```{r 04_pc_variance_plot}
pc |>
  tidy(matrix = "eigenvalues") |>
  filter(PC <= 5) |>
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) 
```

## Save plot

```{r 05_save_plots}

ggsave(plot = biplot, filename = "PCA_on_proteingroups_original.png", width = 20, height = 10, units = "cm", path = here("results"))
```
